A comparative analysis of penalized linear mixed models in structured genetic data

### 2020-10-09 main.tex
Edits to 4 and 5

@@

We begin by comparing the ordinary Lasso, PC-Lasso, and LMM-Lasso as the level of environmental confounding is varied. Estimation accuracy for Lasso, PC-Lasso, and LMM-Lasso was evaluated based on mean squared error (MSE) and broken down into bias and variance components. Figure \ref{fig:mse} presents these results for all three data types in a 1:1 signal to noise (SNR) setting, where $\eta = 0.5$. The top row of Figure \ref{fig:mse} shows that when $\xi = 0$ (i.e., no environmental confounding), Lasso and LMM-Lasso perform comparably. PC-Lasso results in a slight inflation in MSE compared to the other methods in the simulated data sets due to the fact that 10 additional predictors (PCs) have been added to the model, but none of them actually explain any variability in the outcome.

The bottom row of Figure \ref{fig:mse} shows the MSE when $\xi = 0.8$. In this case, both PC-Lasso and LMM-Lasso clearly outperform the ordinary Lasso. This improvement is more substantial in the Empirical and Independent Subpopulations data than in the 1D Linear Admixture data. Qualitatively, this is due to the fact that the Independent Subpopulations and Empirical data sets are characterized by more discrete, nonoverlapping genetic subpopulations, which more directly correspond to the nonoverlapping environmental effects than do the continuous and blended subpopulations of the 1D Linear Admixture data. In other words, a greater proportion of the environmental effect is in the column space of $\boldsymbol{P}_{\bX}$ for the Empirical and Independent Subpopulations data than for the 1D Linear Admixture data. We can quantify the proportion of the environmental component in the column space of $\boldsymbol{P}_{\bX}$ for each data type using the CP metric defined in \eqref{eqn:ratio} of Section \ref{sec:sim_env_conf}. CP $= 0.47, 0.99, 0.61$ for the 1D Linear Admixture, Independent Subpopulations, and Empirical data, respectively. The large CP of the Independent Subpopulations model reflects the extreme confounding and estimation bias for this data, for which the PC-Lasso and LMM-Lasso methods are able to correct. Meanwhile, the less dramatic improvement in MSE for the 1D Linear Admixture data by the PC-Lasso and LMM-Lasso methods confirms what we know from this data's substantially smaller CP value; that there is less confounding and bias which can be corrected for.

@@

\begin{figure}[H]
    \centering
    \includegraphics[scale = 1.1]{figures/beta_mse.png}
    \caption{Estimation accuracy of different methods for different data types in the presence and absence of environmental confounding with coarse subpopulation structure and dichotomous-discordant environmental effects.}
    \label{fig:mse}
\end{figure}
% \anna{talk about karl rohe stuff here - if anywhere}

In addition to MSE, these methods' performance in terms of variable selection accuracy were quantified using the true positive rate (TPR) for the first 50 variables each method allowed into the model. Table \ref{tab:var_sel} displays the mean TPR and corresponding standard deviation in the same settings as those illustrated in Figure \ref{fig:mse}. The TPRs echo the MSE results in that Lasso and LMM-Lasso perform comparably in the absence of environmental confounding across all data types, while the addition of the 10 unpenalized PCs leads to a slight decrease in TPR in this setting for the 1D Linear Admixture and Independent Subpopulations data sets. In the presence of environmental confounding, both LMM-Lasso and PC-Lasso achieve substantial improvements in TPR over Lasso across data types. As with the MSE data, these improvements are more substantial in the Independent Subpopulations and Empirical data sets than in the 1D Linear Admixture data. Again, this is due to the smaller proportion of the environmental effect in the column space of $\boldsymbol{P}_{\bX}$ for the 1D Linear Admixture data, which results in less confounded SNP-phenotype associations to begin with.

@@

%------------------------------%
\subsection{LMM-Lasso outperforms PC-Lasso in the presence of fine subpopulation structure}
%------------------------------%

Given the superior performance of PC-Lasso and LMM-Lasso compared to the ordinary Lasso in the presence of environmental confounding, we next compare these methods in the presence of fine and coarse subpopulation structures. Phenotypes were simulated using a 1:1 SNR, $\eta = 0.5$, and varying levels of environmental confounding, $\xi \in \{0.2, 0.5,0.8\}$.

Figure \ref{fig:big_vs_small} summarizes the results of these simulations for the Independent Subpopulations simulated $\bG$ with environmental confounding generated using a dichotomous-discordant $\bgamma$ structure, though qualitatively similar results were observed for the other data types and $\bgamma$ structures. The y-axis displays the difference in MSE (PC-Lasso - LMM-Lasso) for each simulation and for the various subpopulation structures. The dashed horizontal line at 0 indicates no difference in the performance of PC-Lasso and LMM-Lasso, while differences greater than 0 correspond to smaller MSE for LMM-Lasso, and thus superior estimation accuracy.

In the presence of coarse population structure, there is some overlap in the estimation accuracy of PC-Lasso and LMM-Lasso, as evidenced by the corresponding boxes straddling the 0 line. However, the majority of these boxes fall above 0, indicating the superior performance of LMM-Lasso. In the presence of fine population structure, LMM-Lasso is clearly superior to PC-Lasso in estimation accuracy across all levels of environmental confounding. This is consistent with \citet{hoffman2013correcting}, who showed that principal component adjustment methods represent a low dimensional approximation to linear mixed models. As such, it is reasonable that PC-Lasso, which corrects for population structure using a low dimensional set of surrogate variables, would perform reasonably well in the coarse subpopulation setting, but fall short in the presence of fine structure. 
\begin{figure}[H]
    \centering
    % \includegraphics[scale = 0.9]{figures/mse_diff_subpops}
     \includegraphics[scale = 0.9]{figures/mse_diff_subpops.png}
    \caption{Relative performance of LMM-Lasso, PC-Lasso for fine and coarse subpopulation structures and varying levels of environmental confounding with Independent Subpopulations data, $\eta = 0.5$, and dichotomous-discordant environmental effects.}
    \label{fig:big_vs_small}
\end{figure}

@@

In addition to evaluating PC-Lasso and LMM-Lasso in the presence of coarse and fine subpopulation structure, we also consider distinct patterns of environmental effects. Figure \ref{fig:big_small_gamma} shows the difference in MSE (PC-Lasso - LMM-Lasso) broken down by both subpopulation and coefficient structure. Results are shown for phenotypes generated with $\eta = 0.5$, $\xi = 0.8$, and the Independent Subpopulations simulated $\bG$.

@@

@@

LMM-Lasso-Rakitsch uses a two-step approach. First, the variance components are estimated under the assumption of a null model. Then, the estimated variance components are used to transform the data, and the $\bbeta$ are estimated using coordinate descent. LMM-Lasso-ggmix seeks to improve upon this procedure by re-estimating the variance components and $\bbeta$ iteratively. Although this seems a reasonable approach, we have been unable to find a scenario where LMM-Lasso-ggmix outperforms the LMM-Lasso-Rakitsch in estimation accuracy of $\bbeta$. In order to systematically compare both methods, we carried out simulations where $p = 1000$, $n = \{200, 400, 600, 800, 1000, 1200, 1400\}$, and $\eta = \{0.2, 0.5, 0.8\}$. Figure \ref{fig:eta_beta_mse} shows the MSE for estimation of the parameters $\bbeta$ and $\eta$, in addition to the corresponding computing time required.

LMM-Lasso-Rakitsh and LMM-Lasso-ggmix achieve nearly identical MSE for $\bbeta$ in many scenarios, particularly where $n << p$, as shown in the top panel of Figure \ref{fig:eta_beta_mse}. However, when $n \ge p$ the estimation performance of LMM-ggmix becomes erratic, and in some cases worse than that of LMM-Lasso-Rakitsch. This is most dramatic in the presence of substantial environmental environmental confounding, $\xi = 0.8$, and $n = \{1200, 1400\}$.

@@

Overall these results show that LMM-Lasso-ggmix performs as well as LMM-Lasso-Rakitsch in estimating $\bbeta$ and outperforms LMM-Lasso-Rakitsch in estimating $\eta$ when $n << p$. However, these are also the settings in which LMM-Lasso-ggmix is most computationally inefficient. The bottom panel of Figure \ref{fig:eta_beta_mse} shows the computing time required of each method in the various simulation settings, and shows that when $n << p$, LMM-Lasso-ggmix is up to 60 times slower than LMM-Lasso-Rakitsch. Conversely, in settings where $n \ge p$, LMM-Lasso-ggmix is up to 10 times faster than LMM-Lasso-Rakitsch. However, these are also settings in which LMM-Lasso Rakitsch substantially outperforms LMM-Lasso-ggmix in estimating $\bbeta$ and $\eta$ accurately and consistently.

@@

In this work, we have reviewed the concept of population stratification and explicitly defined an environmental mechanism by which it may confound the genotype-phenotype relationship. We have provided a broad outline of existing methods for ameliorating the effects of these unobserved confounding influences. In addition, we have provided a thorough review of two popular approaches for correcting for population structure in genetic data: PC-Lasso and LMM-Lasso.

@@

We considered two implementations of the LMM-Lasso method: LMM-Lasso-Rakitsch and LMM-Lasso-ggmix, which differ in their variance estimation procedures and computational implementations. Although we found that both methods perform similarly in terms of $\bbeta$ estimation accuracy, LMM-Lasso-ggmix has several drawbacks. Firstly, it suffers from estimation instability in $n \ge p$ settings. Secondly, although LMM-Lasso-Rakitsch and LMM-Lasso-ggmix perform comparably when $n << p$, LMM-Lasso-ggmix requires up to 60 times the computing time of LMM-Lasso-Rakitsch in these settings due to its iterative estimation procedure. The additional time required of LMM-Lasso-ggmix may be worthwhile if the objective of an analysis is to accurately estimate the narrow-sense heritability in $n << p$ settings. However, the LMM-Lasso-Rakitsch method is more robust to varying levels of environmental confounding and sample sizes and is considered by the authors to be more broadly applicable.

@@

As an additional point of interest, recall that we provide qualitative and quantitative explanations of how the degree to which differential environmental exposures may confound the genotype-phenotype relationship is dependent upon its similarity to population structure. While the objective of these explanations, and this work as a whole, is to elucidate the manner in which failing to correct for unobserved environmental effects can bias SNP effect estimates, the result shown in \ref{eqn:partition_e} also highlights the manner in which such unobserved and unmodeled environmental effects could lead to substantial bias in narrow-sense heritability estimatess. A more detailed treatment of narrow-sense heritability estimation is beyond the scope of this paper. However, we hope this emphasizes the broad implications and importance of the subtle distinction between the effect of population structure and unobserved environmental heterogeneity.

Finally, it is worth nothing that even in scenarios where LMM-Lasso outperforms other methods, its estimation accuracy still suffers due to the bias inherent in penalized regression methods. Because of this, bias-reduction techniques for penalized LMMs are an attractive avenue for future research.\\

>>>>

\pb{Just an FYI, you may want not to call this diagram a ``Figure''.  Sometimes you're restricted in terms of the number of figures you can include, and this doesn't really need to be a figure, just a diagram (like an equation).  Same thing for the other DAG}

@@

We begin by comparing the ordinary Lasso, PC-Lasso, and LMM-Lasso as the level of environmental confounding is varied. Estimation accuracy for Lasso, PC-Lasso, and LMM-Lasso was evaluated based on mean squared error (MSE) and broken down into bias and variance components \pb{let's just report MSE (and maybe MSPE too); breakdown blurs ``bias'' concepts a bit}. Figure \ref{fig:mse} presents these results for all three data types in a 1:1 signal to noise (SNR) setting, where $\eta = 0.5$. The top row of Figure \ref{fig:mse} shows that when $\xi = 0$ (i.e., no environmental confounding), Lasso and LMM-Lasso perform comparably. PC-Lasso results in a slight inflation in MSE compared to the other methods in the simulated data sets due to the fact that 10 additional predictors (PCs) have been added to the model, but none of them actually explain any variability in the outcome.

\begin{figure}[H]
    \centering
    \includegraphics[scale = 1.1]{figures/beta_mse.png}
    \caption{Estimation accuracy of different methods for different data types in the presence and absence of environmental confounding with coarse subpopulation structure and dichotomous-discordant environmental effects.}
    \label{fig:mse}
\end{figure}

The bottom row of Figure \ref{fig:mse} shows the MSE when $\xi = 0.8$. In this case, both PC-Lasso and LMM-Lasso clearly outperform the ordinary Lasso. This improvement is more substantial in the Empirical and Independent Subpopulations data than in the 1D Linear Admixture data. This is due to the fact that the Independent Subpopulations and Empirical data sets are characterized by nonoverlapping genetic subpopulations, meaning that a greater proportion of the environmental effect is in the column space of $\boldsymbol{P}_{\bX}$. We can quantify the proportion of the environmental component in the column space of $\boldsymbol{P}_{\bX}$ for each data type using the CP metric defined in \eqref{eqn:ratio} of Section \ref{sec:sim_env_conf}. Here, we have CP $= 0.47, 0.99, 0.61$ for the 1D Linear Admixture, Independent Subpopulations, and Empirical data, respectively. \pb{let's move the 0.61 to the next paragraph} The large CP of the Independent Subpopulations model reflects the extreme confounding and estimation bias for this data, for which the PC-Lasso and LMM-Lasso methods are able to correct. Meanwhile, the less dramatic improvement in MSE for the 1D Linear Admixture data by the PC-Lasso and LMM-Lasso methods confirms what we know from this data's substantially smaller CP value; that there is less confounding and bias which can be corrected for.

@@

% \anna{talk about karl rohe stuff here - if anywhere}

In addition to MSE, these methods' performance was also assessed in terms of variable selection using the true positive rate (TPR) for the first 50 variables each method allowed into the model. Table \ref{tab:var_sel} displays the average TPR for the same settings as in Figure \ref{fig:mse}. The TPRs echo the MSE results in that Lasso and LMM-Lasso perform comparably in the absence of environmental confounding across all data types, while the addition of the 10 unpenalized PCs leads to a slight decrease in TPR in this setting for the 1D Linear Admixture and Independent Subpopulations data sets. In the presence of environmental confounding, both LMM-Lasso and PC-Lasso achieve substantial improvements in TPR over the ordinary Lasso. As with the MSE data, these improvements are more substantial in the Independent Subpopulations and Empirical data sets than in the 1D Linear Admixture data.

@@

%------------------------------%
\subsection{LMM-Lasso outperforms PC-Lasso for finer population structures}
%------------------------------%

Given the superior performance of PC-Lasso and LMM-Lasso compared to the ordinary Lasso in the presence of environmental confounding, we next compare these methods in the presence of fine and coarse subpopulation structures \pb{as defined in section....}. Phenotypes were simulated using a 1:1 SNR, $\eta = 0.5$, and varying levels of environmental confounding, $\xi \in \{0.2, 0.5,0.8\}$.

\begin{figure}[H]
    \centering
    % \includegraphics[scale = 0.9]{figures/mse_diff_subpops}
     \includegraphics[width=0.8\textwidth]{figures/mse_diff_subpops.png}
    \caption{Relative performance of LMM-Lasso, PC-Lasso for fine and coarse subpopulation structures and varying levels of environmental confounding with Independent Subpopulations data, $\eta = 0.5$, and dichotomous-discordant environmental effects. \pb{This figure could maybe be shorter -- takes up a lot of vertical room that isn't really necessary}}
    \label{fig:big_vs_small}
\end{figure}

Figure \ref{fig:big_vs_small} summarizes the results of these simulations for the Independent Subpopulations and a dichotomous-discordant environmental confounding pattern, though qualitatively similar results were observed for other settings. The y-axis displays the difference in MSE (PC-Lasso - LMM-Lasso) for each simulation and for the various subpopulation structures. The dashed horizontal line at 0 indicates no difference in the performance of PC-Lasso and LMM-Lasso, while differences above 0 correspond to LMM-Lasso outperforming PC-Lasso.

Overall, in all of the settings, LMM-Lasso outperformed PC-Lasso, although the difference was considerably more pronounced when the subpopulation structure was fine.  These findings are consistent with \citet{hoffman2013correcting}, who showed that principal component adjustment methods represent a low dimensional approximation to linear mixed models. As such, it is reasonable that PC-Lasso, which corrects for population structure using a low dimensional set of surrogate variables, would perform reasonably well in the coarse subpopulation setting, but fall short in the presence of fine structure.

@@

In addition to evaluating PC-Lasso and LMM-Lasso in the presence of coarse and fine subpopulation structure, we also consider various patterns of environmental effects. Figure \ref{fig:big_small_gamma} shows the difference in MSE (PC-Lasso - LMM-Lasso) broken down by both subpopulation and \pb{``coefficient structure'' is vague; can we describe this better?} coefficient structure. Results are shown for phenotypes generated with $\eta = 0.5$, $\xi = 0.8$, and Independent Subpopulations.

\pb{What do you think about dropping the exponential/dichotomous thing and just going with dichotomous throughout?  Doesn't really seem super important.  Well, we can just keep it in mind and if we need to cut down on space / figures, we can think about it.}

@@

\pb{Not sure I love this figure, as it emphasizes exp vs dichotomous, which isn't really very interesting.  Can we re-panel it so that it emphasizes discordant vs concordant?}

@@

LMM-Lasso-Rakitsch uses a two-step approach. First, the variance components are estimated under the assumption of a null model. Then, the estimated variance components are used to transform the data, and the $\bbeta$ are estimated using coordinate descent. LMM-Lasso-ggmix seeks to improve upon this procedure by re-estimating the variance components and $\bbeta$ iteratively. Although this seems a reasonable approach, we have been unable to find a scenario where LMM-Lasso-ggmix reliably outperforms the LMM-Lasso-Rakitsch in estimation accuracy of $\bbeta$. In order to systematically compare both methods, we carried out simulations where $p = 1000$, $n = \{200, 400, 600, 800, 1000, 1200, 1400\}$, and $\eta = \{0.2, 0.5, 0.8\}$. Figure \ref{fig:eta_beta_mse} shows the MSE for estimation of the parameters $\bbeta$ and $\eta$, in addition to the corresponding computing time required.

LMM-Lasso-Rakitsh and LMM-Lasso-ggmix achieve nearly identical MSE for $\bbeta$ in many scenarios, particularly where $n << p$ \pb{you want $n \ll p$, not $n << p$; this occurs a bunch}, as shown in the top panel of Figure \ref{fig:eta_beta_mse}. However, when $n \ge p$ the estimation performance of LMM-ggmix becomes erratic, and in some cases worse than that of LMM-Lasso-Rakitsch. This is most dramatic in the presence of substantial environmental environmental confounding, $\xi = 0.8$, and $n = \{1200, 1400\}$.

@@

Overall these results show that LMM-Lasso-ggmix performs as well as LMM-Lasso-Rakitsch in estimating $\bbeta$ and outperforms LMM-Lasso-Rakitsch in estimating $\eta$ when $n << p$. However, these are also the settings in which LMM-Lasso-ggmix is least computationally efficient. The bottom panel of Figure \ref{fig:eta_beta_mse} shows the computing time required of each method in the various simulation settings, and shows that when $n << p$, LMM-Lasso-ggmix is up to 60 times slower than LMM-Lasso-Rakitsch. Conversely, in settings where $n \ge p$, LMM-Lasso-ggmix is up to 10 times faster than LMM-Lasso-Rakitsch. However, these are also settings in which LMM-Lasso Rakitsch substantially outperforms LMM-Lasso-ggmix in estimating $\bbeta$ and $\eta$ accurately and consistently.

@@

In this work, we have reviewed the concept of population stratification and explicitly defined an environmental mechanism by which it may confound the genotype-phenotype relationship \pb{not sure what this sentence means; what are you referring to?}. We have provided a broad outline of existing methods for ameliorating the effects of these unobserved confounding influences. In addition, we have provided a thorough review of two popular approaches for correcting for population structure in genetic data: PC-Lasso and LMM-Lasso. \pb{emphasize the \emph{penalized} nature of our work here, which is I think what makes it novel}

@@

We considered two implementations of the LMM-Lasso method: LMM-Lasso-Rakitsch and LMM-Lasso-ggmix, which differ in their variance estimation procedures and computational implementations. Although we found that both methods perform similarly in terms of $\bbeta$ estimation accuracy, LMM-Lasso-ggmix has several drawbacks. Firstly, it suffers from estimation instability in $n \ge p$ settings. Secondly, although LMM-Lasso-Rakitsch and LMM-Lasso-ggmix perform comparably when $n << p$, LMM-Lasso-ggmix requires up to 60 times the computing time of LMM-Lasso-Rakitsch in these settings due to its iterative estimation procedure. The additional time required of LMM-Lasso-ggmix may be worthwhile if the objective of an analysis is to accurately estimate the narrow-sense heritability in $n << p$ settings. However, the LMM-Lasso-Rakitsch method is more robust to varying levels of environmental confounding and sample sizes and thus more broadly applicable.

@@

As an additional point of interest, recall that we provide qualitative and quantitative explanations of how the degree to which differential environmental exposures may confound the genotype-phenotype relationship is dependent upon its similarity to population structure. While the objective of these explanations, and this work as a whole, is to elucidate the manner in which failing to correct for unobserved environmental effects can bias SNP effect estimates, equation \eqref{eqn:partition_e} also highlights the manner in which such unobserved and unmodeled environmental effects could lead to substantial bias in narrow-sense heritability estimatess. A more detailed treatment of narrow-sense heritability estimation is beyond the scope of this paper. However, we hope this emphasizes the broad implications and importance of the subtle distinction between the effect of population structure and unobserved environmental heterogeneity.

Finally, it is worth nothing that even in scenarios where LMM-Lasso outperforms other methods, its estimation accuracy still suffers due to the bias inherent in penalized regression methods. Because of this, bias-reduction techniques for penalized LMMs are a potentially attractive avenue for future research.
